---
title: "Building a Spam Filter"
author: "Abe Ceasar Perez"
date: "April 14, 2022"
output: html_document
---

### Implementing Naive Bayes on a Spam Filter

This project aims to build an algorithm that would filter out spam from SMS messages. Specifically we'll be using the Naive Bayes algorithm from a previously classified dataset to predict the probability of classification and actually classifying an SMS message whether it's spam or not.

In addition to the objectives above, we'll be applying the following concepts in this project:

- Conditional Probability
- Bayes Theorem
- Naive Bayes Algorithm

### Loading the Dataset

The dataset that we'll be loading is authored by Tiago A. Almeida and Jose Maria Gomez Hidalgo which can be downloaded in The UCI Machine Learning Repository. For this project, however, we'll only be using the modified dataset which is linked here: <https://dq-content.s3.amazonaws.com/475/spam.csv>

```{r load_data, echo=TRUE, warning=FALSE}

set.seed(1)

suppressPackageStartupMessages(library(tidyverse)) # loads the packages without warnings

df <- read_csv("spam.csv", show_col_types = FALSE) # loads the dataset without showing the columns 
head(df) # displays the first few rows

```

Based on the initial inspection of the dataset, there are exactly 1,000 messages that we'll be using for classifying and testing the model and 2 variables - one for the classification of the text and another for the actual message itself.

```{r inspect_data, echo=TRUE, warning=FALSE}

dim(df) # checks the number of rows and columns

```

Next, we'll be checking how many of the messages are spam and how many are non-spam. We've determined that there are 150 spam messages and 850 non-spam messages.

```{r inspect_sms_type, echo=TRUE, warning=FALSE}

responses <- df$label # stores all labels into a vector
table(responses) # counts the number of spam/not spam messages

```

### Creating the datasets

For the project to be successful, the goal must align towards maximizing the predictive ability of the model or in context of the problem, enhance the ability of the model to correctly classify new messages. 

In order to do so, we'll be creating the datasets listed below based on our loaded dataset:

- A training set, which we'll use to "train" the computer how to classify messages.
- A cross-validation set, which we'll use to assess how different choices of alpha affect the prediction accuracy
- A test set, which we'll use to test how good the spam filter is with classifying new messages.

For the purposes of this project, we'll be creating proportional samples per label to ensure lower bias in our classification. We'll first start by sampling the indeces of the message starting with spam messages

```{r spam_indeces, echo=TRUE, warning=FALSE}

# gets the spam message indeces to sample the dataset
spam_indeces <- which(responses == "spam") 

# generates random samples for 80% of the sampled indeces
train_spam <- sample(spam_indeces,
                     length(spam_indeces)*0.8) 

# filters out the remaining 20% of the spam indeces as the test/cv indeces
init_spam_test <- spam_indeces[!(spam_indeces %in% train_spam)] 

# generates random samples from the remaining 10% of the sampled indeces
cv_spam <- sample(init_spam_test, 15) 

# separates the remaining 10% of the left indeces as the test data indeces
test_spam <- init_spam_test[!(init_spam_test %in% cv_spam)] 

```

We'll also be doing the same procedure above for non-spam messages:

```{r non_spam_indeces, echo=TRUE, warning=FALSE}

# gets the non-spam message indeces to sample the dataset
ham_indeces <- which(responses == "ham") 

# generates random samples for 80% of the sampled indeces
train_ham <- sample(ham_indeces,
                    length(ham_indeces)*0.8) 

# filters out the remaining 20% of the non-spam indeces as the test/cv indeces
init_ham_test <- ham_indeces[!(ham_indeces %in% train_ham)] 

# generates random samples from the remaining 10% of the sampled indeces
cv_ham <- sample(init_ham_test, 85) 

# separates the remaining 10% of the left indeces as the test data indeces
test_ham <- init_ham_test[!(init_ham_test %in%
                              cv_ham)] 

```

And lastly, we'll be using the generated indeces above to create our datasets. We'll also confirm if the proportion of spam and non-spam messages are appropriate for each of the dataset (which is confirmed to be true).

```{r created_datasets, echo=TRUE, warning=FALSE}

# combines the non-spam and spam training indeces to generate the train dataset
train_df <- df[c(train_ham, train_spam),] 

# combines the non-spam and spam cross validation indeces to generate the cross validation dataset
cv_df <- df[c(cv_ham, cv_spam),] 

# combines the non-spam and spam test indeces to generate the test dataset
test_df <- df[c(test_ham, test_spam),] 

# checks the proportion of spam and non-spam messages for each dataset created
prop.table(table(train_df$label))
prop.table(table(cv_df$label))
prop.table(table(test_df$label))

```

### Cleaning the Data

Since we'll be calculating probabilities for each message, we want to make sure that each word is cleaned and easily recognizable for the algorithm to work. We'll first start by creating a function that would perform the data cleaning for each dataset:

```{r clean_sms, echo=TRUE, warning=FALSE}

message_cleaner <- function(messages){
  
  # removes tab spaces and new lines
  messages <- str_remove_all(messages, "\n|\t") 
  
  # removes punctation marks
  messages <- str_replace_all(messages, "[:punct:]", "") 
  
  # removes numbers/digits
  messages <- str_replace_all(messages, "[:digit:]", "") 
  
  # removes special characters which are not punctuation marks
  messages <- str_replace_all(messages, "[+||||£|$|~|=]", "")
  
   # changes message to lower case
  messages <- str_to_lower(messages)
  
  # removes additional spaces in-between words 
  messages <- str_squish(messages) 
  messages
  
}

```

We'll then apply this function to each of our datasets to finally clean our sms messages.

```{r apply_cleaning, echo=TRUE, warning=FALSE}

# cleans the messages for each created dataset
train_df$sms <- message_cleaner(train_df$sms)
cv_df$sms <- message_cleaner(cv_df$sms)
test_df$sms <- message_cleaner(test_df$sms)

head(train_df)

```

### Creating a Vocabulary from the Training Set

Next, we'll be creating a vocabulary from the training dataset to store our frequency counts for both labels for calculating the probabilities. We'll first create a function to further filter out all valid English words found in each of the datasets that we've created for our vocabulary:

```{r create_vocab, echo=TRUE, warning=FALSE}

# loads the package to extract all possible english words from the Scrabble game
suppressPackageStartupMessages(library(words)) 

df_words <- words # creates a dataframe from the loaded dataset
word_checker <- function(x){
  if(x == "") # checks if the message is empty or not
    x
  else{
    # splits text by spaces
    word_vector <- unlist(str_split(x, pattern=" ")) 
    
    # stores invalid words in a different vector
    invalid_words <- word_vector[word_vector %in% df_words$word == FALSE] 
    
  if(length(invalid_words) > 0){ # checks whether there are invalid words in the vector
    x <- word_vector[match(invalid_words, word_vector)*-1] # removes all invalid words from the vector
    x <- paste(x, collapse = " ") # combines the vector again by spaces after the removal of invalid words
  }
  else
    x
  }
}

train_df$sms <- unlist(map(train_df$sms, word_checker))
cv_df$sms <- unlist(map(cv_df$sms, word_checker)) # cleans the dataset as well for better testing later
test_df$sms <- unlist(map(test_df$sms, word_checker)) # cleans the dataset as well for better testing later

```

From here, we'll be creating a vocabulary from the training dataset in order to calculate the probabilities later.

```{r apply_vocab, echo=TRUE, warning=FALSE}

# extracts all unique words from the training set and sorts it alphabetically
vocabulary <- sort(unique(unlist(map(train_df$sms, function(x) str_split(x, pattern=" "))))) 

# considers only words that contain two letters or more
vocabulary <- vocabulary[str_length(vocabulary) > 1] 
head(vocabulary)

```

### Building the Naive Bayes Algorithm

Now that we have our cleaned messages and vocabulary, we can proceed with calculating the probabilities for each message. In order for the Naive Bayes algorithm to work, we need the following:

- Total number of words in spam messages, non-spam messages, and vocabulary
- Smoothing parameter
- Marginal probabilities
- Word count per class

We'll start first by generating the word counts for spam, non-spam and vocabulary words as well as declaring a smoothing parameter:

```{r marginal_counts, echo=TRUE, warning=FALSE}

# split the cleaned dataset into spam and non-spam
train_df_ham <- subset(train_df, label=="ham")
train_df_spam <- subset(train_df, label=="spam")

# extracts the words in all of the spam messages
Nspam <- length(unlist(map(train_df_spam$sms, function(x) str_split(x, pattern=" ")))) 

# extracts the words in all of the non-spam messages
Nham <- length(unlist(map(train_df_ham$sms, function(x) str_split(x, pattern=" ")))) 

# total number of words in the vocabulary
Nvocabulary <- length(vocabulary) 

# smoothing parameter to account for the non-occurrence of words
alpha <- 1 

# Print counts
print(paste("Total number of spam words: ", Nspam, sep=""))
print(paste("Total number of non-spam words: ", Nham, sep=""))
print(paste("Total number of vocabulary words: ", Nvocabulary, sep=""))

```

Next, we'll calculate the marginal probabilities of a message being spam or not spam based on the training dataset.

```{r probability_per_class, echo=TRUE, warning=FALSE}

# proportion of spam messages
p_spam <- nrow(train_df_spam)/nrow(train_df) 

# proportion of non-spam messages
p_ham <- nrow(train_df_ham)/nrow(train_df) 

print(paste("Probability of a spam message: ", p_spam, sep=""))
print(paste("Probability of a non-spam message: ", p_ham, sep=""))

```

Lastly, we'll be counting the number of occurrences for every possible word for each class or i.e. how many times a word is present in a spam and a non-spam message. The process is illustrated below:

```{r count_per_class, echo=TRUE, warning=FALSE}

count_occurrences <- function(vect){ # function for counting occurrences
  
  # splits the vector (spam or non-spam by spaces)
  split_text <-  unlist(map(vect, function(sms) str_split(sms, pattern=" "))) 
  
  # maps each word in the vocabulary and counts how many time
  word_counts <- unlist(map(vocabulary, function(word) sum(split_text == word))) 
  word_counts
}

spam_occurrences <- count_occurrences(train_df_spam$sms) 
ham_occurrences <- count_occurrences(train_df_ham$sms)

# combines all occurrences into a vocabulary
vocabulary_occurrences <- tibble(vocabulary, spam_occurrences, ham_occurrences) 

head(vocabulary_occurrences) # displays the first five rows

```

Now that we have all the parameters needed, we can now create our classifier that takes in an sms message and outputs a class based on the created algorithm below:

```{r naive_bayes_algorithm, echo=TRUE, warning=FALSE}

spam_ham_classifier <- function(sentence, alpha){ # function for classifying a message with a smoothing parameter
  # splits text
  word_vector <- unique(unlist(str_split(sentence, pattern=" "))) 
  
  # maps the indeces of the split text in the vocabulary
  vocabulary_indeces <- match(word_vector, vocabulary_occurrences$vocabulary) 
  
  # uses the mapped vocabulary indeces to obtain the frequency of the words in a spam message and calculates the probability
  p_words_given_spam <- unlist(map(vocabulary_indeces, function(word_i) (vocabulary_occurrences$spam_occurrences[word_i] + alpha) / (Nspam + alpha * Nvocabulary))) 
  
  # uses the mapped vocabulary indeces to obtain the frequency of the words in a non-spam message and calculates the probability
  p_words_given_ham <- unlist(map(vocabulary_indeces, function(word_i) (vocabulary_occurrences$ham_occurrences[word_i] + alpha) / (Nham + alpha * Nvocabulary))) 
  
  # calculates the final probability of the message being spam
  p_spam_given_words <- p_spam * prod(p_words_given_spam, na.rm = TRUE) 
  
  # calculates the final probability of the message being non-spam
  p_ham_given_words <- p_ham * prod(p_words_given_ham, na.rm=TRUE) 
  
  if(p_spam_given_words > p_ham_given_words) # compares which probability is larger and returns the final classification
    return("spam")
  else
    return("ham")
  
}

```

### Testing the Created Algorithm

After we have created the algorithm, we'll now start by testing it to our datasets. We'll first start by testing it to the train data itself and based on the results below, we're able to obtain a near 100% accuracy, which should be expected given that the model was trained using the train data.

```{r train_accuracy, echo=TRUE, warning=FALSE}

# maps the train messages in the algorithm and sets the smoothing parameter to 1
classifier_predictions <- unlist(map2(train_df$sms, rep(1,length(train_df$sms)), spam_ham_classifier)) 

# combines the actual classes to the predictions made by the algorithm
updated_train_df <- data.frame(cbind(train_df, classifier_predictions)) 

colnames(updated_train_df)[c(1,3)] <- c("actual","predicted") # renames the columns
table(updated_train_df$predicted, updated_train_df$actual) # creates the confusion matrix for predicted and actual classifications

correctly_classified <- sum(unlist(map(c(1,2), function(index) table(updated_train_df$predicted, updated_train_df$actual)[index,index]))) # adds the number of correctly classified messages
total_classified <- nrow(updated_train_df)

train_accuracy <- correctly_classified/total_classified # calculates the accuracy of the model
train_accuracy

```

Earlier, we've arbitrarily set the smoothing parameter to 1, not knowing other values to consider in the algorithm. In order to pick out the best smoothing parameter to use, we'll be mapping a range of alphas using our cross validation set to determine the best alpha to use. 

Based on the results below, the alphas that gave the highest accuracies ranged from 0.3 to 0.9. This would imply that a lower range of alphas are performing well compared to a higher range.

```{r cv_accuracy, echo=TRUE, warning=FALSE}

# for each alpha from 0.1 - 1, run the naive bayes algorithm across each message in the cross validation set.
cv_predictions <- map(seq(0.1,1,0.1), function(parameter) unlist(map2(cv_df$sms, rep(parameter,length(cv_df$sms)), spam_ham_classifier))) 

# calculates the accuracy of the algorithm for each alpha and stores it in a vector
cv_accuracy <- unlist(map(cv_predictions, function(x) {sum(x == cv_df$label)/length(cv_df$label)}))  

# creates a dataframe of the alphas used and the accuracies obtained.
cv_parameter_accuracies <- tibble(alpha = seq(0.1,1,0.1), cv_accuracy) 
cv_parameter_accuracies

```

Similarly, we'll also be performing this in the test data set using the process above. Based on the results below, the best smoothing parameters are 0.1 and 1 in terms of the yielded accuracy - which is a sign that 1 may be the most ideal alpha to use given the strong performance.

```{r test_accuracy, echo=TRUE, warning=FALSE}

# for each alpha from 0.1 - 1, run the naive bayes algorithm across each message in the test set.
test_predictions <- map(seq(0.1,1,0.1), function(parameter) unlist(map2(test_df$sms, rep(parameter,length(test_df$sms)), spam_ham_classifier))) 

# calculates the accuracy of the algorithm for each alpha and stores it in a vector
test_accuracy <- unlist(map(test_predictions, function(x) {sum(x == test_df$label)/length(test_df$label)})) 

# creates a dataframe of the alphas used and the accuracies obtained.
test_parameter_accuracies <- tibble(alpha = seq(0.1,1,0.1), test_accuracy) 
test_parameter_accuracies

```

<br>
<br>